---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4)

ggplot(subset(pf, !is.na(gender)), aes(age, friend_count)) +
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```

```{r}
library(dplyr)
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n=n()) %>%
  ungroup() %>%
  arrange(age)
head(pf.fc_by_age_gender, 20)
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
ggplot(pf.fc_by_age_gender, aes(age, median_friend_count)) +
  geom_line(aes(color=gender))
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
# install.packages('reshape2')
# library(reshape2)
library("reshape2", lib.loc="~/Anaconda3/envs/py37pipped/Lib/R/library")

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')
```

```{r}
head(pf.fc_by_age_gender.wide)
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}
ggplot(pf.fc_by_age_gender.wide, aes(female, male)) +
  geom_point() +
  geom_hline(yintercept = 1)
```
```{r}

```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
ggplot(pf.fc_by_age_gender.wide, aes(age, female / male)) +
  geom_line() +
  geom_hline(yintercept = 1, alpha = .3, linetype = 2)
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
pf$year_joined <- floor(2014 - (pf$tenure / 365))
```
```{r}
table(pf$year_joined)
```
```{r}
pf$year_joined.bucket <- cut(pf$year_joined,
                             breaks = c(2004, 2009, 2011, 2012, 2014))
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA = 'ifany')

ggplot(subset(pf, !is.na(gender)), aes(age, friend_count)) +
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```
```{r}
ggplot(subset(pf, !is.na(year_joined.bucket)), aes(age, friend_count)) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```
```{r}
ggplot(subset(pf, !is.na(year_joined.bucket)), aes(age, friend_count)) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) +
  geom_smooth()
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}

```

***

### Friending Rate
Notes:

```{r friending rate}
with(subset(pf, tenure >= 1), summary( friend_count / tenure ))
```


***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}

ggplot(subset(pf, !is.na(year_joined.bucket) & tenure >= 1), aes(age, friendships_initiated / tenure)) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) +
  geom_smooth()
```
```{r}
ggplot(subset(pf, tenure >= 1), aes(tenure, friendships_initiated / tenure)) +
         geom_line(aes(color = year_joined.bucket))
```


```{r}
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 100)) +
  geom_line(aes(color = year_joined.bucket))
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```
```{r}
ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:
```{r}
yo <- read.csv('yogurt.csv')
str(yo)

# Change the id from an int to a factor
yo$id <- factor(yo$id)
str(yo)
```

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
ggplot(yo, aes(price)) +
  geom_histogram(binwidth = 10, fill = 'orange')
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
summary(yo)
length(unique(yo$price))
table(yo$price)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
yo$allpurchases = transform(yo, allpurchases=strawberry + blueberry +
                              pina.colada + plain + mixed.berry)$allpurchases
```
```{r}
# alternate 1
yo <- transform(yo, allpurchases=strawberry + blueberry +
                              pina.colada + plain + mixed.berry)
```
```{r}
qplot(data = yo, x = allpurchases, binwidth=1)
```
```{r}
ggplot(yo, aes(time, price)) +
  geom_point()
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
# set seed for reproducible results
set.seed(4530)
sample.ids <- sample(levels(yo$id), 16)

ggplot(aes(x = time, y = price),
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap(~id) +
  geom_line() +
  geom_point(aes(size = allpurchases), pch = 1)

```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:
```{r}
library("GGally", lib.loc="~/Anaconda3/envs/py37pipped/Lib/R/library")
theme_set(theme_minimal(20))

# set the seed for reproducible results
set.seed(1836)
# long version
pf_subset <- pf[, c(2:15)]
# short version
#pf_subset = pf[, c('age', 'dob_year', 'dob_month', 'gender', 'tenure')]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000),])
```

***

### Even More Variables
Notes:
```{r}
nci <- read.table('nci.tsv')

#changing the colnames to produce a nicer plot
colnames(nci) <- c(1:64)
```
```{r}
# melt the data to long format
library("reshape2", lib.loc="~/Anaconda3/envs/py37pipped/Lib/R/library")
nci.long.samp <- melt(as.matrix(nci[1:200, ] ))
names(nci.long.samp) <- c('gene', 'case', 'value')
head(nci.long.samp)

# make the heat map
ggplot(aes(y = gene, x = case, fill = value),
       data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c('blue','red'))(100))
```

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

