---
title: "EDA_Udacity_R"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
Lesson 3
=======================================================

## Reading in Data
```{r}
pf <- read.csv('pseudo_facebook.tsv', sep='\t')
names(pf)
```

## Histogram of User' Birthdays
```{r}
#install.packages('ggplot2')
library(ggplot2)

names(pf)
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks=1:31)
```

## Faceting
```{r}
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol=3)
```

## Friend Count
```{r}
qplot(data = pf, x = friend_count)
```

## Limiting the Axes
```{r}
qplot(x = friend_count, data = pf, xlim = c(0, 1000))

qplot(x = friend_count, data = pf) +
  scale_x_continuous(limits = c(0, 1000))
```

## Adjusting the bin width
```{r}
qplot(x = friend_count, data = pf, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks=seq(0, 1000, 50)) +
  facet_wrap(~gender)
```


## Omitting 
```{r}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000), breaks=seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

## Statistics 'by' Gender
```{r}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

## Tenure ~months
```{r}
qplot(x = tenure, data = pf, binwidth = 30,
      color = I('black'), fill = I('#099DD9'))
```


## Tenure ~years
```{r}
qplot(x = tenure/365, data = pf, binwidth = .25,
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
  
```

## labeling Plots
```{r}
qplot(x = tenure/365, data = pf, binwidth = .25, xlab = 'Number of years using Facebook',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7))
  
```

## User Ages
```{r}
qplot(x = pf$age, data= pf, binwidth = 10,
      xlab = 'Age of Facebook User',
      ylab = 'Number of users in sample',
      color = I('green'), fill = I('#F79420'))+
  scale_x_continuous(breaks = seq(5, 120, 20), limits = c(20,120))
```
```{r}
ggplot(aes(x = age), data = pf) +
  geom_histogram(binwidth = 1, fill = '#5760AB') +
  scale_x_continuous(breaks = seq(0, 113, 5))
```

## Transforming Data
```{r}
qplot(x = friend_count, data = pf)

summary(pf$friend_count)

summary(log10(pf$friend_count + 1))

summary(sqrt(pf$friend_count))
```

## Tester for function
```{r}
qplot(x = pf$friend_count, data = pf) + scale_x_log10()
```

## Mulitple plots
```{r}
# define individual plots
library(gridExtra) # import gridExtra lib

p1 = qplot(x = pf$friend_count, data = pf) 


p2 = qplot(x = pf$friend_count, data = pf) + scale_x_log10()


p3 = qplot(x = pf$friend_count, data = pf) + scale_x_sqrt()
# arrange plots in grid
grid.arrange(p1, p2, p3,  ncol=1)
```

## Transforming Data Alternate Solution
## Use scales!
```{r}
p1 <- ggplot(aes(x = pf$friend_count), data =pf) + geom_histogram()

p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()

grid.arrange(p1, p2, p3, ncol=1)
```
 

## Frequency Polygons
```{r}
qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(0, 1000), breaks=seq(0, 1000, 50))
```


## Which gender makes more likes?(presolution)
```{r}
qplot(x = www_likes, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Likes Made',
      ylab = 'Proportion of users with that like count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(5, 100), breaks=seq(5, 100, 5))
```


## likes on the web
```{r}
by(pf$www_likes, pf$gender, sum)
```

## Box Plots
```{r}
qplot(x = gender, y = friend_count,
      data = subset (pf, !is.na(gender)),
      geom = 'boxplot')
```



## Box Plots altered
```{r}
qplot(x = gender, y = friend_count,
      data = subset (pf, !is.na(gender)),
      geom = 'boxplot') + 
  scale_y_discrete(limits = c(0, 1000))
```
## Box Plots alternate
```{r}
qplot(x = gender, y = friend_count,
      data = subset (pf, !is.na(gender)),
      geom = 'boxplot', ylim = c(0, 1000))
```
## Box Plots, Quartiles, and Friend Requests
```{r}
qplot(x = gender, y = friend_count,
      data = subset (pf, !is.na(gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim = c(0, 250))

by(pf$friend_count, pf$gender, summary)
```

## Getting logical counts
```{r}
summary(pf$mobile_likes)

summary(pf$mobile_likes > 0)

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)
```


what percent of checkin was mobile?
```{r}
summary(pf$mobile_check_in)[2]/sum(summary(pf$mobile_check_in))
```
-----------------------

# Lesson 4

## Loading Diamonds data set

```{r}
library(ggplot2) #must load the ggplot package first
data(diamonds) #loads the diamonds data set since it comes with the ggplot package
```

## Q1
```{r}
summary(diamonds)
table(unlist(lapply(diamonds, class)))
```

## Q2 price histogram
```{r}
qplot(x = diamonds$price, data = diamonds)
```

## Q3
```{r}
summary(diamonds$price)
```


## Q4
```{r}
summary(diamonds$price < 500)
summary(diamonds$price < 250)
summary(diamonds$price >= 15000)

```

## Q5 cheaper diamonds
```{r}
qplot(x = diamonds$price,
      data = diamonds,
      #xlim = c(0, 15000),
      binwidth = 10,
      ylim = c(0,350)) +
  scale_x_continuous(
    limits=c(0,15000),
    breaks=seq(0,15000,3000)
  )
ggsave(filename = "quiz4CheaperDiamonds.png")
# note an extension must be provided above
```

## q6
```{r}
qplot(x = diamonds$price,  data = diamonds, binwidth=10) +
  facet_wrap(~cut)
```

## q7
```{r}
by(diamonds$price, diamonds$cut, max)
```

## q8
```{r}
qplot(x = price, data = diamonds) + facet_wrap(~cut, scales="free_y")
```

## q9
```{r}
qplot(x = diamonds$price/diamonds$carat, data = diamonds, binwidth=.01) + scale_x_log10() +
  facet_wrap(~cut, scales="free_y")
```


## q10
```{r}
qplot(x = gender, y = friend_count,
      data = subset (pf, !is.na(gender)),
      geom = 'boxplot')
```
## q11 a, b
```{r}
# IQR(subset(diamonds, color='D')$price)
by(diamonds$price, diamonds$color, summary)
IQR(subset(diamonds, color=='D')$price) 
IQR(subset(diamonds, color=='J')$price) 
```
# q11 c, d
```{r}

# IQR(subset(diamonds, color='D')$price)
```

## q12 
