---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:
```{r}
library(ggplot2)
pf <- read.delim('pseudo_facebook.tsv')
```

```{r Scatterplots qplot}
qplot(x = age, y = friend_count, data = pf)
```
```{r Scatterplots}
ggplot(aes(x = age, y = friend_count), data = pf)
```

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) + geom_point()
```
***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax with xlim}
# geom_point() layer necessay in r syntax to create scatterplots
ggplot(aes(x = age, y = friend_count), data = pf) + geom_point() +
  xlim(13, 90)

summary(pf$age)

```

***

### Overplotting
Notes:

```{r Overplotting}

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) +
  xlim(13, 90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(height = 0)) +
  xlim(13, 90) +
  coord_trans(y = "sqrt")
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = pf$friendships_initiated), data = pf) +
  geom_jitter( alpha = 1/10) 

```

***

### Overplotting and Domain Knowledge
Notes:

***
```{r}
library(dplyr)
```

### Conditional Means
Notes:

```{r Conditional Means}
age_groups <- group_by(pf, age)
pf.fc_by_age<- summarise(age_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age) # arrange by age, ascending
head(pf.fc_by_age)
```


```{r}
# %>% pass value from left side into first argument of right side
pf.fc_by_age <- pf %>%
  group_by(age) %>%
         summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age)

head(pf.fc_by_age, 20)
```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x=age, y=friend_count_mean), data = pf.fc_by_age) +
  geom_line()
```

```{r Conditional Means Plot shorthand}
ggplot(pf.fc_by_age, aes(age, friend_count_mean)) +
  geom_line()
```
***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(pf, aes(age, friend_count)) +
  coord_cartesian(xlim = c(13, 70), ylim = c(0, 1000)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
  # coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1), linetype = 'dashed', color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .5), linetype = 'solid', color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9), linetype = 'dashed', color = 'blue')
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor(pf$age, pf$friend_count)
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(   subset(pf, age <= 70, select=c(age, friend_count))              , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(pf, aes(pf$www_likes_received, pf$likes_received)) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')
```
```{r}
cor(pf$www_likes_received, pf$likes_received)
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}

```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
# install.packages('alr3')
library(alr3)
data("Mitchell")
?Mitchell
```

Create your plot!

```{r Temp vs Month}
ggplot(Mitchell, aes(Month, Temp)) +
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
cor(Mitchell$Month, Mitchell$Temp)
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(Mitchell, aes(Month, Temp)) +
  geom_point() +
  scale_x_continuous(breaks=1:12, limit=c(1,12))
```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(pf.fc_by_age, aes(age, friend_count_mean)) +
  geom_line()

head(pf.fc_by_age, 10)
pf.fc_by_age[17:19, ]
```

***

### Age with Months Means

```{r Age with Months Means}
pf$age_with_months <- pf$age + (1 - pf$dob_month / 12)
```




Programming Assignment
```{r Programming Assignment}
#pf.fc_by_age_months <-
# %>% pass value from left side into first argument of right side
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
         summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age_with_months)

head(pf.fc_by_age_months, 20)
```
### 24.

```{r}
ggplot(subset(pf.fc_by_age_months, age_with_months < 71 ), aes(age_with_months, friend_count_mean)) +
  geom_line()
```

***

### Noise in Conditional Means
```{r Noise in Conditional Means}
```

***

### 25 Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
library(gridExtra)
p1 <- ggplot(subset(pf.fc_by_age, age< 71 ), aes(age, friend_count_mean)) +
  geom_line() +
  geom_smooth()
p2 <- ggplot(subset(pf.fc_by_age_months, age_with_months < 71 ), aes(age_with_months, friend_count_mean)) +
  geom_line() +
  geom_smooth()
p3 <- ggplot(subset(pf, age < 71 ), aes(round(age/5)*5, friend_count)) +
  geom_line(stat = 'summary', fun.y = mean)

grid.arrange(p2, p1, p3, ncol = 1)

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

