---
title: "Red Wine Quality EDA by Riley Shea"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(knitr)
library(dplyr)
library(grid)
```
```{r}
install.packages('GGally')
install.packages('scales')
install.packages('memisc')
install.packages('lattice')
install.packages('MASS')
install.packages('car')
install.packages('reshape')
install.packages('plyr')

library(gridExtra)
library(GGally)
library(scales)
library(memisc)

```


```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('winequalityreds.csv', sep = ',')
# drop line number column 'X'
wine <- select(wine, -c(X))

```
This report explores a dataset containing chemical attributes and quality
scores for approximately 1600 samples of red wine.

# Univariate Plots Section
```{r}
# Dimensions of df
dim(wine)
# list the the structure of df
str(wine)
# some quick stats on the columns of df
summary(wine)
```
Our dataset consists of 12 variables with ~1600 observations.

The 'X' variable is merely the row number, so it was dropped after import.
The quality is a median of >= 3 expert's ratings

```{r}
wine$badfso <- factor(wine$free.sulfur.dioxide > 50)
```


```{r Quality histogram}
ggplot(aes(quality), data = wine) +
  geom_histogram(binwidth = 1, fill = '#5760AB') +
  ggtitle('Quality') + 
  scale_x_continuous(breaks = seq(3, 8, 1), limits = c(2, 9))

```

```{r}
ggplot(aes(quality), data=wine) +
  geom_histogram(binwidth = .08, fill = 'brown') +
  ggtitle('Quality log10') + 
  scale_x_log10()

```
This dataset lacks categorical variables and the input variables are fairly
granular. Couple of noteworth exceptions

```{r Fixed Acidity Histogram}
ggplot(aes(fixed.acidity), data = wine) +
  geom_histogram(binwidth = .1, fill = '#5760AB') +
  ggtitle('Fixed Acidity') + 
  scale_x_continuous(breaks = seq(5, 15, 1), limits = c(5, 15))
```

```{r Volatile Acidity Histogram}
ggplot(aes(volatile.acidity), data = wine) +
  geom_histogram(binwidth = .05, fill = '#5760AB') +
  ggtitle('Fixed Acidity') + 
  scale_x_continuous(breaks = seq(0.1, 1.3, .1), limits = c(.1, 1.3)) 
```
```{r Citric Acid Histogram}
ggplot(aes(citric.acid), data = wine) +
  geom_histogram(binwidth = .01, fill = '#5760AB') +
  ggtitle('Citric Acid') +  
  scale_x_continuous(breaks = seq(0.0, 0.8, .1), limits = c(-.1, .8)) 
```

```{r Residual Sugar Histogram}
ggplot(aes(residual.sugar), data = wine) +
  geom_histogram(binwidth = .05, fill = '#5760AB') +
  ggtitle('Residual Sugar') +
  scale_x_log10(breaks = seq(1, 8, 1), limits = c(1, 8)) 
```
```{r chlorides Histogram}
ggplot(aes(chlorides), data = wine) +
  geom_histogram(binwidth = .01, fill = '#5760AB') +
  ggtitle('Chlorides') + 
  scale_x_continuous(breaks = seq(0, .2, .02), limits = c(0.01, .2)) 
```

```{r Free Sulfur Dioxide}
ggplot(aes(free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1, fill = '#5760AB') +
  ggtitle('Free Sulfur Dioxide') + 
  scale_x_continuous(breaks = seq(0, 50, 5), limits = c(0, 50)) 
```
```{r Total Sulfur Dioxide}
ggplot(aes(total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1, fill = '#5760AB') +
  ggtitle('Total Sulfur Dioxide') +
  scale_x_continuous(breaks = seq(0, 160, 10), limits = c(0, 160)) 
```
```{r density}
ggplot(aes(density), data = wine) +
  geom_histogram(binwidth = .001, fill = '#5760AB') +
  ggtitle('density') +
  scale_x_continuous(breaks = seq(0.990, 1.04, .002), limits = c(.990, 1.004)) 
```
```{r pH}
ggplot(aes(pH), data = wine) +
  geom_histogram(binwidth = .1, fill = '#5760AB') +
  ggtitle('pH') +
  scale_x_continuous(breaks = seq(2.8, 3.8, .4), limits = c(2.8, 3.8)) 
```

```{r Sulphates}
ggplot(aes(sulphates), data = wine) +
  geom_histogram(binwidth = .05, fill = '#5760AB') +
  ggtitle('Sulphates') +
  scale_x_continuous(breaks = seq(.25, 1.25, .25), limits = c(.25, 1.25)) 
```
```{r Alcohol}
ggplot(aes(alcohol), data = wine) +
  geom_histogram(binwidth = .1, fill = '#5760AB') +
  ggtitle('Alcohol') + 
  scale_x_continuous(breaks = seq(9, 13, 1), limits = c(9, 13)) 
```
> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
# qplot(x=df$fixed.acidity, data=df, bins = 100)
# qplot(x=df$volatile.acidity, data=df, bins = 100)
# qplot(x=df$citric.acid, data=df, bins = 100)
# qplot(x=df$residual.sugar, data=df, bins = 100)
# qplot(x=df$chlorides,data=df, bins = 100)
# qplot(x=df$free.sulfur.dioxide, data=df, bins = 100)
# qplot(x=df$total.sulfur.dioxide, data=df, bins = 100)
# qplot(x=df$density, data=df, bins = 100)
# qplot(x=pH, data=df, bins = 100)
# qplot(x=sulphates, data=df, bins = 100)
# qplot(x=alcohol, data=df, bins = 100)
# qplot(x=quality, data=df, bins = 100)


```

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

[//]: # "Comment"

### What is the structure of your dataset?

There are 15,999 wine samples in the dataset with 12 features(fixed acidity,
volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide,
total sulfur dioxide, density, pH, sulphates, alchohol, and quality).
Quality is an output variable of integer type.  The remaining 11 variables are
input variables of numeric type.

Other observations:

* There are much more normal wines than there are 'poor' or 'excellent' wines.
The vast majority range from 5-to-7
* The median quality is a 6, and the max quality score is an 8.
* All wines have a pH far below 7, making their pH 'basic'.


### What is/are the main feature(s) of interest in your dataset?

From this dataset I am most interested in the 'quality' variable and how it 
interacts with:

* volatile acidity

> "volatile acidity: the amount of acetic acid in wine, which at too high of
levels can lead to an unpleasant, vinegar taste"

* citric acid

> "citric acid: found in small quantities, citric acid can add 'freshness'
and flavor to wines"

* free sulfur oxide

> "...at free SO2 concentrations     over 50 ppm, SO2 becomes evident in the
nose and taste of wine"
 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The 'residual sugar' variable might also have been of interest since it is an
indicator of sweetness.  Seeing if there was a perceived quality difference for
sweet or dry red wines might have been interesting.
Unfortunately all the wines in the dataset fell far below the 45 g/L(dm^3).

### Did you create any new variables from existing variables in the dataset?

I would've liked to create some categorical variables for input variables
with thresholds that define a wine characteristic (sweet/dry, acidic/basic, 
) unfortunately the wine values weren't split between the two thresholds. 



### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Only tidying I did was to remove the 'X' field as it was redudant with R's own
row names.


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r Geom point: Volatile acidity echo=FALSE, Bivariate_Plots}
ggplot(wine, aes(volatile.acidity, quality)) +
  scale_x_continuous(lim = c(0, quantile(wine$volatile.acidity, 0.99))) +
  scale_y_continuous(lim = c(0, quantile(wine$quality, 0.99))) +
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21)
```

It looks like there might be a pattern here with this geom_point graph but it's
a little hard to discern.  Lets try using quality as a factor and use that to
generate a boxplot instead.

```{r Bivariate Volatile acidity boxplots}
ggplot(wine, aes( factor(quality), volatile.acidity, color=factor(quality))) +
  geom_boxplot(outlier.colour = "red") +
  stat_summary(fun.y = mean,
               geom = 'point',
               shape = 8) 

```

This makes it much easier to see the pattern.  As quality rating increases both
the median and mean volatile acidity decreases. This is true for each rating
with the exception of quality rating 8 for which there was very little data.

```{r quality-volatile.acidity correlation}
cor(wine$quality, wine$volatile.acidity)
```

This confirms the plot findings showing a medium-strength negative correlation
between quality and volatile acidity.


```{r Bivariate Citric acid boxplots}
ggplot(wine, aes(factor(quality), citric.acid, color = factor(quality))) +
  geom_boxplot(outlier.colour = "red") +
  stat_summary(fun.y = mean,
  geom = 'point',
  shape = 8) 
```

Here we see another clear pattern.  In this case as citric acid increases the
factor increases too.

```{r quality-}
cor(wine$quality, wine$citric.acid)
```

This seems to indicate a small positive correlation between citric acid and
quality

```{r Bivariate Free Sulpher dioxide boxplots}
ggplot(wine, aes( factor(quality), free.sulfur.dioxide,
                  color=colorRampPalette(c("blue", "red")))) +
  geom_boxplot(outlier.colour = "red" ) +
  stat_summary(fun.y = mean,
               geom = 'point',
               shape = 8) 

```

This graph is perhaps the most interesting so far. Free sulphur dioxide seems
positively correlated with quality up until rank 5, at which point it trends
downward.  This seems to coincide with wineQualityInfo description which stated:

> "...at free SO2 concentrations     over 50 ppm, SO2 becomes evident in the
nose and taste of wine"

However since the relation doesn't look linear the correlation coefficient may
not give us the full picture.

```{r}
cor(wine$quality, wine$free.sulfur.dioxide)
```

This is a case where the relationship coefficient seems to have failed us.

```{r Creation wine.by_quality dataframe}
wine.by_quality <- wine %>%
  group_by(quality) %>%
  summarise(mean_volatile.acidity = mean(volatile.acidity),
            median_volatile.acidity = median(volatile.acidity),
            n=n()) %>%
  ungroup() %>%
  arrange(quality)
head(wine.by_quality, 20)
```
```{r ggpairs graph(all bivariate graphs)}
# No need sample with a seed for this size of sample
ggpairs(wine, 
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

We saw three different types of relationships:

* medium-strength, negative linear between quality and volatile acidity
* low-strength, positive linear between quality and citric acid
* non-linear relationship 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

From glancing over the ggpairs output it was interesting to note the
high-strength positive correlation between citric acid and fixed acidity,
in constrast to it's high strength negative correlation with volatile acidity.


### What was the strongest relationship you found?

The strongest relationship from the points of interest was volitile acidity
to quality.

From The entire dataset a high-strength negative correlation of -0.683 for 
'fixed acidity' and 'pH' seemed to be the strongest correlation I saw.

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r citric.acid by volatile.acidity and quality echo=FALSE, Multivariate_Plots}

# install and load the RColorBrewer package
library("RColorBrewer", lib.loc="~/Anaconda3/envs/py37pipped/Lib/R/library")

ggplot(aes(x = volatile.acidity, y = citric.acid, colour = factor(quality)), data = wine) + 
  geom_point(alpha = 1, size = 2) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous( limits = c(0.15, 1.25),
     breaks = seq(0.15, 1.5, .25)) + 
  scale_y_continuous( limits = c(-.1, .8),
     breaks = seq(0, .8, .1)) + 
  ggtitle('citric.acid by volatile.acidity and quality')
```
```{r}
ggplot(aes(x = free.sulfur.dioxide, y = badfso,
           colour = factor(quality)), data = wine) + 
  geom_point(alpha = 1, size = 1.5) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Above 50ppm', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  # scale_x_continuous( limits = c(0.15, 1.25),
  #    breaks = seq(0.15, 1.5, .25)) + 
  # scale_y_continuous( limits = c(-.1, .8),
  #    breaks = seq(0, .8, .1)) + 
  ggtitle('Free sulfur oxide > 50 ppm')
```
```{r}
ggplot(aes(x = free.sulfur.dioxide, y = badfso,
           colour = factor(quality)), data = wine) + 
  geom_point(alpha = 1, size = 1.5) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Above 50ppm', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  # scale_x_continuous( limits = c(0.15, 1.25),
  #    breaks = seq(0.15, 1.5, .25)) + 
  # scale_y_continuous( limits = c(-.1, .8),
  #    breaks = seq(0, .8, .1)) + 
  ggtitle('Free sulfur oxide > 50 ppm')
```
```{r}
ggplot(wine, aes(free.sulfur.dioxide, total.sulfur.dioxide)) +
geom_point(aes(color = factor(quality)), stat = 'summary', fun.y = median)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!