---
title: "Red Wine Quality EDA by Riley Shea"
output:
  html_document: 
    highlight: pygments
editor_options:
  chunk_output_type: console
---
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# These lines may need to be uncommented if the packages aren't installed
# install.packages('GGally')
# install.packages('scales')
# install.packages('memisc')
# install.packages('lattice')
# install.packages('MASS')
# install.packages('car')
# install.packages('reshape')
# install.packages('plyr')
library(RColorBrewer)
library(grid)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(ggplot2)
library(knitr)
library(dplyr)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load the Data
wine <- read.csv('winequalityreds.csv', sep = ',')
# drop line number column 'X'
wine <- select(wine,-c(X))
```

This report explores a data set containing chemical attributes and quality
scores for approximately 1600 samples of red wine.

# Univariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Dimensions of df
dim(wine)
# list the the structure of df
str(wine)
# some quick stats on the columns of df
summary(wine)
```

Our data set consists of 12 variables with ~1600 observations.

The 'X' variable is merely the row number, so it was dropped after import.
The quality is a median of >= 3 expert's ratings

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine$badfso <- factor(wine$free.sulfur.dioxide > 50)
```



```{r echo=FALSE, message=FALSE, warning=FALSE}
# use ggplot2 to create histogram
ggplot(aes(quality), data = wine) +
  geom_histogram(binwidth = 1, fill = '#5760AB') +
  ggtitle('Quality') +
  scale_x_continuous(breaks = seq(3, 8, 1), limits = c(2, 9))
```

Unfortunately the quality is somewhat imbalanced. Perhaps applying a log10
transform will help.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(quality), data = wine) +
  geom_histogram(binwidth = 1, fill = 'brown') +
  ggtitle('Quality log10') +
  scale_y_log10()
```

This transformed quality histogram looks much more normal.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(fixed.acidity), data = wine) +
  geom_histogram(binwidth = .5, fill = '#5760AB') +
  ggtitle('Fixed Acidity Continuous') +
  scale_x_continuous(breaks = seq(5, 15, 1), limits = c(4, 15))
```

This fixed acidity histogram looks distribution looks fairly normal without a
log transform.  Lets try one anyway.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(fixed.acidity), data = wine) +
  geom_histogram(binwidth = .04, fill = '#5760AB') +
  ggtitle('Fixed Acidity Log10') +
  scale_x_log10()
```

Looks a bit more normal, but not significantly so.

```{r}
qplot(data = wine, x = volatile.acidity, bins = 50)
```

If we omit the low counts to the right this volatile acidity histogram looks
pretty normalized.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(volatile.acidity), data = wine) +
  geom_histogram(binwidth = .05, fill = '#5760AB') +
  ggtitle('Volatile Acidity') +
  scale_x_continuous(breaks = seq(0.1, 1.3, .1), limits = c(.1, 1.3)) 
```

Yes, omitting those low counts seems to help the shape.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(citric.acid), data = wine) +
  geom_histogram(binwidth = .01, fill = '#5760AB') +
  ggtitle('Citric Acid') +
  scale_x_continuous(breaks = seq(0.0, 0.8, .1), limits = c(-.1, .8)) 
```

This long tailed histogram of citric acid seems to have a large bin of samples
with values with less than .01 grams/L of citric acid. My guess would be that
citric acid might be an additive that some winemakers choose to omit when
crafting their wine.

Also see a smaller spike ~.5 which could potentially be a result of less precise
measurements or maybe a value that some winemakers aim to have in their wine.

```{r}
ggplot(aes(wine$citric.acid ), data = wine) +
  geom_histogram(binwidth = 0.04, fill = '#5760AB') +
  ggtitle('Citric Acid') +
  scale_x_log10(breaks=seq(.0, 1, .05), limits= c(.2, .9)) 
  
```

Transforming this Citric Acid histogram seems to confirm our suspicions, showing
a bi modal graph with a couple peaks at .25, and .5.

```{r}
qplot(data= wine, x = residual.sugar, bins = 50)
```

Initially this residual sugar looks somewhat long tailed, but limiting the
x axis should give us a more normal shape

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(residual.sugar), data = wine) +
  geom_histogram(binwidth = .25, fill = '#5760AB') +
  ggtitle('Residual Sugar') +
  scale_x_continuous(breaks = seq(1, 4, .25), limits = c(1, 3.75)) 
```

That looks a bit better

```{r}
qplot(data = wine, x = chlorides, bins = 50)
```

This chlorides histogram looks like it would be normal if we remove some of
these potential outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(chlorides), data = wine) +
  geom_histogram(binwidth = .01, fill = '#5760AB') +
  ggtitle('Chlorides') +
  scale_x_continuous(breaks = seq(0, .16, .01), limits = c(0.02, .14)) 
```

This looks limit helps confirm a normal shape.


```{r}
qplot(data = wine, x = free.sulfur.dioxide, bins = 20)
```

This free sulfur dioxide appears long-tailed so we'll try transforming it to
better understand it's shape.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = .2, fill = '#5760AB') +
  ggtitle('Free Sulfur Dioxide scale_x_log10') +
  scale_x_log10(limits = c(1,100))
```

This scale helps use see it's uni modal distribution.


```{r}
qplot(data = wine, x = total.sulfur.dioxide, bins = 20)
```

Just as with free sulfur dioxide we see a long tailed histogram. Lets try
another log_x_10 layer.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = .2, fill = '#5760AB') +
  ggtitle('Total Sulfur Dioxide scale_x_log10') +
  scale_x_log10()
```

Once again applying this scale_x_log10 helps us see the normal, uni modal shape.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(density), data = wine) +
  geom_histogram(binwidth = .001, fill = '#5760AB') +
  ggtitle('density') +
  scale_x_continuous(breaks = seq(0.990, 1.04, .002), limits = c(.990, 1.004)) 
```

This density histogram already shows a normal shape, doesn't seem to need a
log10 scale.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(pH), data = wine) +
  geom_histogram(binwidth = .1, fill = '#5760AB') +
  ggtitle('pH') +
  scale_x_continuous(breaks = seq(2.8, 3.8, .4), limits = c(2.8, 3.8)) 
```

'pH' shows a normal shape for it's histogram.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(sulphates), data = wine) +
  geom_histogram(binwidth = .05, fill = '#5760AB') +
  ggtitle('Sulfates') +
  scale_x_continuous(breaks = seq(.25, 1.25, .25), limits = c(.25, 1.25)) 
```

Sulfates looks fairly normal here.

```{r}
qplot(data = wine, x = alcohol, binwidth = 1) +
  scale_x_continuous(breaks = seq(7, 16, 1))
```

Could argue this alcohol histogram is long tailed or normal I think.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(alcohol), data = wine) +
  geom_histogram(binwidth = .05, fill = '#5760AB') +
  ggtitle('Alcohol scale_x_sqrt') + 
  scale_x_sqrt()
```

Applying a square root scale doesn't seem to change the shape.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(alcohol), data = wine) +
  geom_histogram(binwidth = .01, fill = '#5760AB') +
  ggtitle('Alcohol') + 
  scale_x_log10(breaks = seq(8, 14, .5), limits=c(8.5, 13.5))
```

Looks to be a somewhat bi modal, possibly multi-modal shape here.  

### What is the structure of your dataset?

There are 15,999 wine samples in the data set with 12 features(fixed acidity,
volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide,
total sulfur dioxide, density, pH, sulfates, alcohol, and quality).
Quality is an output variable of integer type.  The remaining 11 variables are
input variables of numeric type.

Other observations:

* There are much more normal wines than there are 'poor' or 'excellent' wines.
The vast majority range from 5-to-7
* The median quality is a 6, and the max quality score is an 8.
* All wines have a pH far below 7(~3-to-4), making their pH 'basic'.


### What is/are the main feature(s) of interest in your dataset?

From this data set I am most interested in the 'quality' variable and how it 
interacts with:

* volatile acidity

> "volatile acidity: the amount of acetic acid in wine, which at too high of
levels can lead to an unpleasant, vinegar taste"

* citric acid

> "citric acid: found in small quantities, citric acid can add 'freshness'
and flavor to wines"

* free sulfur dioxide

> "...at free SO2 concentrations  over 50 ppm, SO2 becomes evident in the
nose and taste of wine"
 

### What other features in the dataset do you think will help support your
### investigation into your feature(s) of interest?

The 'residual sugar' variable might also have been of interest since it is an
indicator of sweetness.  Seeing if there was a perceived quality difference for
sweet or dry red wines might have been interesting.
Unfortunately all the wines in the data set fell far below the 45 g/L(dm^3), 
so all observations were by definition dry wines.

### Did you create any new variables from existing variables in the dataset?

I would've liked to create some categorical variables for input variables
with thresholds that define a wine characteristic (sweet/dry, acidic/basic, 
) unfortunately the wine values weren't split between the two thresholds. 



### Of the features you investigated, were there any unusual distributions? 
### Any operations on the data to tidy, adjust, or change the form 
### of the data? If so, why did you do this?

Only change I made was to remove the 'X' field as it was redundant with R's own
row names.  The source described the data as being tidy already.


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(volatile.acidity, quality)) +
  scale_x_continuous(lim = c(0, quantile(wine$volatile.acidity, 0.99))) +
  scale_y_continuous(lim = c(0, quantile(wine$quality, 0.99))) +
  geom_point(
  fill = I('#F79420'),
  color = I('black'),
  shape = 21
  )
```

It looks like there might be a pattern here with this geom_point graph but it's
a little hard to discern.  Lets try using quality as a factor and use that to
generate a box plot instead.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(factor(quality), volatile.acidity, color = factor(quality))) +
  geom_boxplot(outlier.colour = "red") +
  stat_summary(fun.y = mean,
  geom = 'point',
  shape = 8)
  
```

This makes it much easier to see the pattern.  As quality rating increases both
the median and mean volatile acidity decreases. This is true for each rating
with the exception of quality rating 8 for which there was very little data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$quality, wine$volatile.acidity)
```

This confirms the plot findings showing a medium-strength negative correlation
between quality and volatile acidity.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(factor(quality), citric.acid, color = factor(quality))) +
  geom_boxplot(outlier.colour = "red") +
  stat_summary(fun.y = mean,
               geom = 'point',
               shape = 8) 
```

Here we see another clear pattern.  In this case as citric acid increases the
factor increases too.

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$quality, wine$citric.acid)
```

This seems to indicate a small positive correlation between citric acid and
quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(factor(quality), free.sulfur.dioxide,
                 color = factor(quality))) +
                 geom_boxplot(outlier.colour = "red") +
                 stat_summary(fun.y = mean,
                 geom = 'point',
                 shape = 8)
                 
```

This graph is perhaps the most interesting so far. Free sulfur dioxide seems
positively correlated with quality up until rank 5, at which point it trends
downward.  This seems to coincide with wineQualityInfo.txt description
which stated:

> "...at free SO2 concentrations over 50 ppm, SO2 becomes evident in the
nose and taste of wine"

However since the relation doesn't look linear the correlation coefficient may
not give us the full picture.

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$quality, wine$free.sulfur.dioxide)
```

This is a case where the relationship coefficient seems to have failed us.

The quality attribute takes on so few values it effectively acts like an
ordered factor.  With that in mind we'll add a factor attribute based on quality
below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine$fQuality <- factor(wine$quality)
```

Quick look over our variables of interest to see if we missed anything
important.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# No need sample with a seed for this size of sample
ggpairs(
subset(
wine,
select = c(
"volatile.acidity",
"citric.acid",
"free.sulfur.dioxide",
"fQuality"
)
),
lower = list(continuous = wrap("points", shape = I('.'))),
upper = list(combo = wrap("box", outlier.shape = I('.'))),
axisLabels = 'internal'
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(data = wine,  x = volatile.acidity) +
  facet_wrap( ~ fQuality, scales="free_y") 
```

Unfortunately this faceted plot doesn't seem very useful since the
classes(quality scores) aren't balanced.

# Bivariate Analysis

We saw three different types of relationships:

* medium-strength, negative linear between quality and volatile acidity
* low-strength, positive linear between quality and citric acid
* non-linear relationship between free sulfur dioxide and quality

### Did you observe any interesting relationships between the other features 
### (not the main feature(s) of interest)?

From glancing over the ggpairs output it was interesting to note the
high-strength positive correlation between citric acid and fixed acidity,
in contrast to it's high strength negative correlation with volatile acidity.

### What was the strongest relationship you found?

The strongest relationship from the points of interest was volatile acidity
to quality.

From The entire data set a high-strength negative correlation of -0.683 for 
'fixed acidity' and 'pH' seemed to be the strongest correlation I saw.

```{r}
cor(wine$fixed.acidity, wine$pH)
```


# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(
x = volatile.acidity,
y = citric.acid,
colour = factor(quality)
), data = wine) +
geom_point(alpha = 1, size = 2) +
scale_color_brewer(type = 'div',
guide = guide_legend(
title = 'quality',
reverse = T,
override.aes = list(alpha = 1, size = 2)
)) +
scale_x_continuous(limits = c(0.15, 1.25),
breaks = seq(0.15, 1.5, .25)) +
scale_y_continuous(limits = c(-.1, .8),
breaks = seq(0, .8, .1)) +
ggtitle('citric.acid by volatile.acidity and quality')
```

Presumably, citric acid would account for some of the volatile acidity.
However their effect on quality is the opposite.  As seen in this scatter plot
their is a cluster of high quality samples with low volatile acidity and high
citric acid values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(
  x = free.sulfur.dioxide,
  y = badfso,
  colour = factor(quality)
  ), data = wine) +
  geom_point(
  aes(color = factor(quality)),
  stat = 'summary',
  fun.y = median,
  size = 5,
  alpha = .8
  ) +
  ggtitle('Free sulfur dioxide > 50 ppm')
```

This confirms what we suspected earlier. The free sulfur dioxide increases
quality up preserving freshness, however if the levels become too high it
affects the taste quality. None of the samples with free S02 over 50ppm
achieved quality of 8.  

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(free.sulfur.dioxide, total.sulfur.dioxide)) +
  geom_point(aes(color = factor(quality)), stat = 'summary', fun.y = median) +
  scale_color_brewer(type = 'seq',
  guide = guide_legend(
  title = 'quality',
  reverse = T,
  override.aes = list(alpha = 1, size = 2)
  )) +
  scale_x_continuous(limits = c(1, 58),
  breaks = seq(1, 58, 10)) +
  scale_y_continuous(limits = c(5, 125),
  breaks = seq(5, 125, 10)) +
  ggtitle('free S02 by total S02 and quality')
```

Seem to see greater dispersion between samples as total S02 and free SO2
increase. At lower values they seem more tightly clustered together.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(
  x = free.sulfur.dioxide,
  y = citric.acid,
  colour = factor(quality)
  ),
  data = wine) +
  geom_line(
  aes(color = factor(quality)),
  stat = 'summary',
  fun.y = median,
  size = 1,
  alpha = .5
  ) +
  scale_color_brewer(
  type = 'qual',
  guide = guide_legend(
  title = 'Quality ',
  reverse = T,
  override.aes = list(alpha = 1, size = 2)
  )
  ) +
  ggtitle('Citric Acid by Free SO2 and Quality')
```

Not seeing a clear relationship between between these variables.  This makes
sense because citric acid and Free sulfur dioxide don't seem to be
related chemically.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the  
### investigation. Were there features that strengthened each other in terms of 
### looking at your feature(s) of interest?

The first plot in multi-variate analysis I had expected to see citric acid
go up as volatile acidity went up.  I didn't really see this though. It did
reinforce what we saw in our bi variate box plots with high citric acid and low
volatile acidity being beneficial to quality

In the second multi variate scatter plot we saw that none of the top quality
samples had a free sulfur dioxide over 50ppm.

In our third multivariate scatter plot we didn't see a clear relationship
between all three variables.  So a bit of a dead end there.

The final multi-variate plot used a line plot, attempting to find a relationship
between citric acid, free sulfur dioxide, and quality.  Was unable to identify
any distinct trends.

### Were there any interesting or surprising interactions between features?

Had trouble finding notable multivariate interactions between our values of
interest.  Nothing surprising that went against the statements made in the
description of values provided in the data set description text file.

### OPTIONAL: Did you create any models with your dataset? 

No models were created.

------

# Final Plots and Summary

Ultimately, it seems the bi variate plots we created provided the most relevant
insights.

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(factor(quality), volatile.acidity, color = factor(quality))) +
  geom_boxplot(outlier.colour = "red") +
  stat_summary(fun.y = mean,
  geom = 'point',
  shape = 8) +
  labs(
    title = "Volatile Acidity by Quality Boxplot",
    x = "Quality rating (score between 0 and 10)", 
    y = "volatile acidity (acetic acid - g / dm^3)",
    colour = "Quality"
  )
```

### Description One

Here we see a distinct trend, as volatile acidity goes down, the quality rating
goes up.  Keeping the volatile acidity low in wine looks to be a key component
of receiving a high quality rating.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(factor(quality), citric.acid, color = factor(quality))) +
  geom_boxplot(outlier.colour = "red") +
  stat_summary(fun.y = mean,
  geom = 'point',
  shape = 8) +
  labs(
  title = "Citric Acid by Quality Boxplot",
  x = "Quality rating (score between 0 and 10)",
  y = "Citric acid (g / dm^3)",
  colour = "Quality"
  )
```

### Description Two

Once again we see another clear trend.  A higher citric acid level in wine seems
to directly affect it's quality rating. In this case we see the relationship is
positive and linear.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine, aes(factor(quality), free.sulfur.dioxide,
                 color = factor(quality))) +
                 geom_boxplot(outlier.colour = "red") +
                 stat_summary(fun.y = mean,
                 geom = 'point',
                 shape = 8) +
                 labs(
                 title = "Free Sulfur Dioxide by Quality Boxplot",
                 x = "Quality rating (score between 0 and 10)",
                 y = "Free Sulfur Dioxide (mg / dm^3)",
                 colour = "Quality"
                 )
```

### Description Three

Here we see a non-linear relationship between Free Sulfur Dioxide and Quality.
This was in my mind the most interesting observation.  It seems that Free Sulfur
oxide acts as a 'double-edged blade'.  It helps prevent microbial growth and
oxidation in wine, key factors affecting quality. The downside being that if the
levels become too high it affects taste and nose(smell).

This is made very evident in this box plot figure.  The lowest scores show the
lowest levels of FSO2.  As a result these wines likely experienced cases of
oxidation and/or microbial growth adversely affecting their quality.

The medium quality scores(5 and 6), had the highest FSO2 concentrations. It
seems most likely that in this case that oxidation and microbial growth were
largely prevented, helping these wine to score higher quality scores.  FS02's
impact on taste and nose at high levels likely prevented them from achieving the
highest quality scores.

The highest quality scores showed lower levels of FS02 than their mid-tier
counterparts.  Likely that at the highest quality level alternate methods to
prevent oxidation and microbial growth were used.  Presumably a more expensive
process than simply keeping FS02 high.

------

# Reflection

Ultimately this data set seemed to not be great for multivariate analysis. Had
the data set contained information like grape types, wine brand,
wine selling price, etc I feel much more interesting analysis could have been
done.  Regardless, we did see some interesting bi variate relationships between
our input variables of interest and their impact on the output
variable 'quality'.

If I were to do further analysis I'd try to wrangle a bigger set, with some of
the aforementioned attributes.  Then I could try to determine things
like 'highest quality wine brands' or 'best value wines'.
